---
title: "Final Project"
author: "STAT 4610/5610 Spring 2026"
format: pdf
editor: visual
date: 2026-03-30
---

## Final Project: Applying Various Methods to Answer Questions of Interest

**Due: 11:59PM Friday, April 24 on Canvas as a knitted pdf file of your team's Quarto document**

1.  You will individually fit **at least three statistical learning models** to answer your individual question of interest based on your team's chosen data set. You will individually explain what the relevant problem context is (Q1), what methods you used and their results (Q2), how those results answer the question of interest (Q3), and what implications (recommendations, next step actions, etc.) the answers have (Q3). In addition, you will comment on some ethical aspect of the problem.
2.  **Each individual must use at least one unique method**, i.e., not a method used by any of your teammates. For example, if team member 1 fit a cubic spline, BART, and a feedforward neural network to answer their individual research question and member 2 used a cubic spline, BART, and random forest, and member 3 used BART, natural spline, and lasso, then member 4 could not use a feedforward NN or a random forest or both natural spline and lasso.
3.  Teams will propose and answer a team research question that should incorporate the individual contributions. In other words, teams will combine findings from the individual components when answering their team question. Each individual must contribute to the team part and state what their contribution was. No stated contribution by individuals means no team points for that individual.
4.  Team answers should follow the Q1-Q2-Q3 workflow, including a reflection about an ethical principal used or some other ethical aspect of the team solution. Teams must also produce a plot that helps to tell the narrative of the data. (Individuals should also include plots in their individual parts, but it's not strictly required.)

#### What teams need to do first (Final Project Proposal)

1.  Choose a dataset that has enough interesting predictor X variables so that you can have four individual sub-research questions and one overall team research question. You are encouraged to gain practice doing unsupervised learning (PCA, K-means or hierarchical clustering) if that makes sense for your data set and your research questions.
2.  Brainstorm various research questions. Each individual needs their own sub-research question to answer. Think about how the answers to the sub-research questions could inform the answer to the overall team research question.

Overall research question: What factors influence outcomes in traffic stops?
Corva's sub-research question: Can driver characteristics like age, sex, race, predict whether the driver is given a warning, citation, or arrested? I will use K-Nearest Neighbors, QDA, BART

#### What individuals need to do

1.  Given your team's data set and your sub-research question, fit at least three statistical learning models. For each model fit, you should also play around with different tuning parameters, likely using CV.
2.  Explain Q1, Q2, and Q3.
3.  Comment on any ethical implications of this work.
4.  Highly recommended: Create visualizations of your models or just of the best model.

#### What teams need to do

1.  Answer a team research question.
2.  Explain Q1, Q2, and Q3.

## Corva's individual part

#### K-Nearest Neighbors

K-Nearest Neighbors is selected because n is very large (\~14,000), and p is relatively small. No shape about the decision boundary is assumed.


```{r}
#K-Nearest Neighbors
#will try a variety of K

data2 <- data %>%
  filter(!is.na(subject_age), !is.na(subject_race)) %>%
  select(outcome, subject_age, subject_race, subject_sex)

knn.Y <- data2$outcome
knn.X <- model.matrix(~ subject_age + subject_race + subject_sex -1, data = data2)

#get rows
set.seed(123)
crows <- sample(nrow(knn.X), size = floor(0.8 * nrow(knn.X)))

ctrain.X <- knn.X[crows, ]
ctest.X  <- knn.X[-crows, ]

ctrain.X <- scale(ctrain.X)
ctest.X  <- scale(ctest.X, 
                  center = attr(ctrain.X, "scaled:center"),
                  scale  = attr(ctrain.X, "scaled:scale"))

ctrain.Y <- knn.Y[crows]
ctest.Y  <- knn.Y[-crows]

ctrain <- data2[crows, ]
ctest  <- data2[-crows, ]

#adapting code from KNN.R in canvas

error.rate.cv.knn <- rep(NA, 15)
for (k in 1:15) {
  out <- knn.cv(
    train = ctrain.X,
    cl = ctrain.Y,
    k= k
  )
  
  error.rate.cv.knn[k] <- mean(out != ctrain.Y)
}

plot(error.rate.cv.knn, type = "b")
error.rate.cv.knn
```

Using cross-validation to pick the best k for KNN, it appears by inspection of the testing error rate that the best k is 10.

```{r}
# fitting the final model and finding error rate

final.knn.c <- knn(ctrain.X, ctest.X, ctrain.Y, k=10)


# error rate
error.knn.c <- mean(final.knn.c != ctest.Y)

table(Predicted = final.knn.c, Actual = ctest.Y)
```

Using K-Nearest Neighbors, with k selected with cross-validation, error rate was around 23% in predicting whether a driver was given a warning, citation, or arrested using just driver sex, age, and race. However, even after scaling variables and adjusting k, the model defaults to the null and only predicts the outcome of citation. It appears that the three predictors of driver characteristics, age, race, and sex, don't provide much predictive power for outcome.


#### Multinomial Logistic Regression

The next model, multinomial logistic regression, is selected because it might be able to pick out patterns relating between driver characteristics and outcome across all of the data points, instead of just looking at potentially noisy neighbors.

```{r}
set.seed(123)

multi.log.c <- cv.glmnet(x = ctrain.X, y = ctrain.Y, family = "multinomial", type.measure = "class")
multi.lambda <- multi.log.c$lambda.min

multi.log.pred <- predict(multi.log.c, ctest.X, multi.lambda, type="class")
table(Predicted = multi.log.pred, Actual = ctest.Y)

mean(multi.log.pred == ctest.Y)
```

In fitting the logistic regression, cross-validation was performed for $\lambda$, which controls how closely the model is allowed to follow the training data. It appears that the model ends up just predicting the outcome citation, which is the most frequent class, for all data points. Again, it seems that the three predictors of driver characteristics, age, race, and sex, don't provide much predictive power for outcome.

#### Random Forests

Random forests are selected for the next model. Random forests decorrelate trees by using a random subset of predictors. As opposed to K-Nearest-Neighbors or logistic regression, random forests give more consistent predictions by averaging many trees.  These earlier models defaulted to the null model, and random forests are able to explore more complex relationships that may have been missed. 

```{r}
set.seed(123)
data2$outcome <- as.factor(data2$outcome)


#cross validate for node size and size of number of predictors considered

mtry.valuesF <- 1:3
oob.errors <- numeric(length(mtry.valuesF))

for (j in seq_along(mtry.valuesF)) {
  rf <- randomForest(
    outcome ~ subject_age + subject_race + subject_sex,
    data = data2,
    mtry = mtry.valuesF[j],
    ntree = 200
  )

  oob.errors[j] <- tail(rf$err.rate[, "OOB"], 1)
}

rforestbestvals <- which.min(oob.errors) 
rforestbestvals

#fit tuned model and predict and find validated test error


randomFTree.tuned <- randomForest(outcome~subject_age+subject_race+subject_sex,
                                  data=ctrain,
                                  mtry=1,
                                  ntree=200,
                                  importance=TRUE)

randomFTree.pred <- predict(randomFTree.tuned,newdata=ctest)
table(Predicted = randomFTree.pred, Actual = ctest$outcome)
```

In fitting the random forests model, the parameter $mtry$ was tuned, which controls how much data the model looks at in each split. Along with K-Nearest Neighbors and Logistic Regression, the model ends up just predicting the outcome citation, which is the most frequent class, for all data points. It seems that the three predictors of driver characteristics, age, race, and sex, don't provide much predictive power for outcome.

2.  Explain Q1, Q2, and Q3.

How well do driver characteristics like age, sex, and race, inform outcome of traffic stops? This sub-question informs the main research question by analyzing how driver characteristics, and nothing else about the traffic stop, impact whether the driver is given a warning, citation, or is arrested.

Three statistical learning models were applied to data about the driver age, race, and sex. K-Nearest Neighbors was selected because no shape about any decision boundary was assumed. Next, multinomial logistic regression was fitted because it looks at all of the data points, not just neighbors. Finally, a random forests model was fitted to the data because it is able to explore more complex relationships between the three predictors, and because it produces more reliable results by averaging trees. However, all three models, even after tuning parameters, defaulted to just predicting the most common outcome, citation. 

All three of the models, even after tuning, just predicted the most common outcome and were not able to find any meaningful relationships between the predictors and the outcome. It appears that, only using driver characteristics of age, race, and sex, that outcome cannot meaningfully be predicted. Further analysis is needed - these predictors may be significant in the context of other predictors, for example, district or time of day, however, do not have predictive power by themselves. 

3.  Comment on any ethical implications of this work.

Even though these models did not find any predictive power of the three driver characteristics, it is important not to assume that age, race, and sex don't have any affect on traffic stop outcomes. Traffic stop outcomes are highly contextual and driver characteristics may still have influence on them when considering more predictors.




Q1: What factors influence outcomes in traffic stops, given police traffic stop data in Hartford, CT from October 2013-September 2016? Governments and police departments are interested in this question in order to more fairly enforce outcomes of traffic stops. More specifically, are outcomes of traffic stops more influenced by the drivers, or the officers conducting the traffic stops? Are specific situations of traffic stops, for example, stops that are late at night on certain days, more likely to have different outcomes? Data was obtained from the Stanford Open Policing Project, which collects and standardizes records of traffic stops across the United States.

3.  Comment on any ethical implications of this work.
4.  Create at least one visualization that helps tell the overall narrative of the data.

#### Individual contributions

Each individual must comment on their contributions to the team research question. Only individuals who contribute to the team section (and state their contributions) will get points for the team section.

#### **Some intended outcomes from this final project:**

-   You will individually get practice fitting statistical learning models
-   You will consolidate your learning of statistical learning methods to answer research questions of interest to you
-   You will gain experience working on a data science project as a team
-   You can post your final project on a personal portfolio (like a GitHub page) to show an example of your work to potential employers
