---
title: "Lab3"
author: "STAT 4610-5610: Spring 2026"
format: pdf
editor: visual
date: 2026-03-23
---

## Lab3: Support Vector Machines

**Due: 11:59PM Friday, March 27 on Canvas as a knitted pdf file of your team's Quarto document**

1.  You will individually fit a support vector machine, describe what it is doing, and report its misclassification rate, sensitivity, and specificity.
2.  You will interpret this model and make a prediction in your individual section.
3.  As a team, you will create a support vector machine (SVM) and compete against other teams for the best test performance. You will also create a team plot that summarizes the performance of the individual SVMs and your team's SVM.

### Instructions for Lab3

In Lab0 (and Lab1), the professor created a generating model for Y variables taking on 0 or 1 values based on the X1 and X2 inputs. You came up with the backstory for what Y, X1, and X2 were and why it was important to use X1 and X2 to predict Y.

In this lab, you will apply support vector machines to continue your analyses for how to best predict Y from X1 and X2. You should continue to use the Q1 qualitative context you developed in Lab0 or Lab1 (or you can develop a new one if you want).

Each individual will fit at least one SVM on a training dataset and then evaluate how well the SVM classifies Y based on a test dataset. Individuals will make a prediction given (x1, x2) and then interpret their prediction and make a recommendation for action. Unlike in Lab0 and Lab1, you will not describe the whole workflow of Q1, Q2, and Q3 for this "project". You will report on the misclassification rate, sensitivity, and specificity for your SVM and make a prediction given (x1, x2). You will comment on how the prediction differs from your prediction in Lab1 and, given this new prediction, describe what actions (Q3) you recommend given your Q2 results.

#### Generating Model

We have a logistic regression generating model. Given $x_1 \in [0,1]$ and $x_2 \in [0,1], Y \sim Ber(p)$, where $p$ is related to $x_1$ and $x_2$ through the logistic link function: $\log(\frac{p}{(1-p)}) = x_1 -2x_2 -2{x_1}^{2} + {x_2}^{2} + 3{x_1}{x_2}$, where $\log$ is the natural log, sometimes written as $\ln$.

The code for this is below.

```{r}
library(class)
suppressPackageStartupMessages(library(tidyverse))
install.packages("e1071")
library(e1071)
install.packages("caret")
library(caret)
install.packages(tidyr)
library(tidyr)
install.packages(dplyr)
library(dplyr)
install.packages(ggplot2)
library(ggplot2)
```

```{r}
#Generative model
set.seed(200) #setting a random seed so that we can
#reproduce everything exactly if we want to

generate_y <- function(x1,x2) { #two input parameters to generate the output y
  logit <- x1 -2*x2 -2*x1^2 + x2^2 + 3*x1*x2 
  p <- exp(logit)/(1+exp(logit)) #apply the inverse logit function
  y <- rbinom(1,1,p) #y becomes a 0 (with prob 1-p) or a 1 with probability p
}
```

#### Training dataset

We are going to use our generating model to create a training dataset of 1000 predictors (x1, x2), and then 1000 outcomes. Then we plot all three variables to see what our training data looks like.

```{r}
# Generate a training dataset with 1000 points
set.seed(321)
n = 1000
X1 <- runif(n,0,1)
X2 <- runif(n,0,1)

#I'm going to use a for loop to generate 1000 y's
Y <- rep(0,n) #initializing my Y to be a vector of 0's
for (i in 1:n) {
  Y[i] <- generate_y(X1[i],X2[i])
}

sum(Y) #How many 0's and 1's were predicted? When n is very large
# about 48% of the Y's should be 1's. 

training <- cbind(X1,X2,Y) #combining all of my variables into a training dataset
ggplot(data=training, aes(x=X1, y=X2, color=Y)) +
  geom_point()

testing <- data.frame(testing)
training <- data.frame(training)
colnames(testing) <- colnames(training)
training$Y <- as.factor(training$Y)
testing$Y <- as.factor(testing$Y)
```

How well will various SVMs predict Y given new x1 and x2 values?

#### Test datasets

Each individual will generate a test set of 100 predictors (x1, x2) and outcomes (y) that we will use as our "ground truth". The test dataset should be the same for each team member. Use seed=323.

```{r}
# Create a testing dataset using seed=323

set.seed(323)
n = 100

testX1 <- runif(n,0,1)
testX2 <- runif(n,0,1)

testY <- rep(0,n) #initializing my Y to be a vector of 0's
for (i in 1:n) {
  testY[i] <- generate_y(testX1[i],testX2[i])
}

testing <- cbind(testX1,testX2,testY) #combining all of my variables into a training dataset
ggplot(data=testing, aes(x=testX1, y=testX2, color=testY)) +
  geom_point()
```

#### What individuals need to do (20 pts)

1.  Given the training set (seed=321), fit a ***different*** SVM for each team member. Choose from the options below:
    1.  linear SVM (this will be a support vector classifier, i.e., a flat hyperplane \[actually just a line\])

## Corva Individual Section

```{r}
set.seed(123)
svctune <- tune(svm , 
                Y~., 
                data = data.frame(training) , 
                kernel = "linear", 
                ranges = list(cost = c(0.001 , 
                                       0.01, 
                                       0.1, 
                                       1, 
                                       5, 
                                       10, 
                                       100)),
                scale=TRUE)
summary(svctune)

bestsvc <- svctune$best.model
summary(bestsvc)

svcpred <- predict(bestsvc, testing)
scvmatrix <- confusionMatrix(data=svcpred,
                             reference=testing$Y)
scvmatrix

#my point is (0.25, 0.25)

cpoint <- data.frame(X1=0.25, X2=0.25)
cpointpred <- predict(bestsvc, cpoint)
as.numeric(cpointpred)
```

The Support Vector Classifier classifies points by finding a hyperplane that best separates two classes. It adds a soft margin between classes that allows some observations to be on the wrong side of the margin.

After predicting on the test set, the Support Vector Classifier has a sensitivity of 1, specificity of 0, and a 49% misclassification rate. This model is probably not a good fit: the model is giving up and is only predicting the most frequent class which is 0. The data is probably not linearly separable, and the plot of X1 vs X2 confirms this. In this problem, Y was whether or not a borrower defaulted on their loan (default=1,no default=0), given their age(X1) and income (X2). The specific prediction is for an age 25 borrower (X1=0.25) who earns 25,000 dollars a year (X2=0.25). This SVC model predicts that the borrower will default on their loan, which is consistent with the results of Lab1. I would recommend not lending to this 25 year old borrower who earns 25,000 dollars a year since all of the models used in Lab1 in addition to this SVC model predicted that he would default on his loan. 


### Parker Individual Section

```{r}
# Create the training dataset as above using seed=314
set.seed(314)
n = 1000
x1 <- runif(n,0,1)
x2 <- runif(n,0,1)

y <- rep(0,n)
for (i in 1:n) {
  y[i] <- generate_y(x1[i],x2[i])
}

train <- cbind(x1,x2,y) 
ggplot(data=training, aes(x=x1, y=x2, color=y)) +
  geom_point()

# Create a testing dataset using seed=323

set.seed(323)
n = 100

xt1 <- runif(n,0,1)
xt2 <- runif(n,0,1)

yt <- rep(0,n)
for (i in 1:n) {
  yt[i] <- generate_y(xt1[i],xt2[i])
}

test <- cbind(xt1,xt2,yt)
test <- as.data.frame(test)

train$y <- as.numeric(as.character(train$y))
train$y[train$y == 0] <- -1

test$yt <- as.numeric(as.character(test$yt))
test$yt[test$yt == 0] <- -1
```
#### 1,2

```{r}
#Fit and cross validate for paramteters of svm

set.seed(123)
train <- data.frame(train)
train$y <- as.factor(train$y)
test$yt <- as.factor(test$yt)

svm_tune <- tune(svm,y~.,data=train,kernel="polynomial",degree=2,
                 ranges = list(cost=c(0.1,1,10,100),
                               gamma = c(0.5,1,2,3)))
svm_tune$best.parameters

svm_fit <- svm(y~., data=train, kernel="polynomial",degree=2,cost=10,gamma=0.5)
```
#### 3

Unlike the maximal margin classifier, the support vector classifier allows for some points to be on the wrong side of the decision boundary. The cost parameter indicates how tolerant we are to points being on the wrong side of the boundary. The cost parameter determines how large the margin is-the area that allows for points to be on the wrong side of the decision. A large c means that we are more tolerant to violations of the decision boundary, and the margin will widen. All the points on one side of the boundary will be given the same prediction. The support vector machine allows us to deal with non linear decision boundaries by expanding the feature space using kernels. The support vector machine combines the support vector classifier with a non-linear kernel.

```{r}
#Create confusion matrix
test <-data.frame(test)
svm_preds <- predict(svm_tune$best.model,newdata=test)

table(preds=svm_preds[1:100],truth=test$yt)
```
The SVM with degree 2 polynomial had a sensitivity of 74%, specificity of 26%, and an overall error rate of 50%. 

#### 5

```{r}
#Make prediction for point (0.25, 0.75)
point <- data.frame(x1=0.25,x2=0.75)
point_pred <- predict(svm_tune$best.model,newdata=point)
point_pred[1]
```
Compared to lab 1, this prediction is different than all of the other models that we compared in lab 1, which is interesting. In the textbook it mentions that logistic regression and the support vector classifier are related, but that logistic regression performs better when the data is mixed together, and the support vector classifier works better when the data is grouped together. The fact that all of the other models in lab 1 predicted 1, I would be suspicious about the svm model. The data in our example is pretty mixed, so it would probably make more sense to use a logistic regression model over a support vector classifier. I imagine that this work is for a financial institution trying to predict fraud, and prediction accuracy is the most important metric, specifically the sensitivity of the model. Although, the svm model had a different prediction compared to the models from lab 1, the svm model has the highest sensitivity of rate of all the models. Since the financial institution is most concerned with the prediction accuracy and sensitivity rate, I think that the svm model may make the most sense in this context.  

##Averey Individual Section

    2.  SVM with an interaction between X1 and X2

Training dataset

We are going to use our generating model to create a training dataset of 1000 predictors (x1, x2), and then 1000 outcomes. Then we plot all three variables to see what our training data looks like.

```{r}
# Generate a training dataset with 1000 points
set.seed(321)
n = 1000
X1 <- runif(n,0,1)
X2 <- runif(n,0,1)

#I'm going to use a for loop to generate 1000 y's
Y <- rep(0,n) #initializing my Y to be a vector of 0's
for (i in 1:n) {
  Y[i] <- generate_y(X1[i],X2[i])
}

sum(Y) #How many 0's and 1's were predicted? When n is very large
# about 48% of the Y's should be 1's. 

training <- cbind(X1,X2,Y) #combining all of my variables into a training dataset
ggplot(data=training, aes(x=X1, y=X2, color=Y)) +
  geom_point()
```
How well will various SVMs predict Y given new x1 and x2 values?

Test datasets

Each individual will generate a test set of 100 predictors (x1, x2) and outcomes (y) that we will use as our "ground truth". The test dataset should be the same for each team member. Use seed=323.

```{r}
# Create the training dataset as above using seed=321
# Create a testing dataset using seed=323
set.seed(323)
n = 100

testX1 <- runif(n,0,1)
testX2 <- runif(n,0,1)

testY <- rep(0,n) #initializing my Y to be a vector of 0's
for (i in 1:n) {
  testY[i] <- generate_y(X1[i],X2[i])
}

testing <- cbind(X1,X2,Y) #combining all of my variables into a training dataset
ggplot(data=testing, aes(x=X1, y=X2, color=Y)) +
  geom_point()
```

```{r}
#view(testing)
```
What individuals need to do (20 pts)

Given the training set (seed=321), fit a different SVM for each team member. Choose from the options below:

linear SVM (this will be a support vector classifier, i.e., a flat hyperplane [actually just a line])

SVM with an interaction between X1 and X2

SVM with degree 2 polynomials

SVM with degree 3 polynomials

Use cross validation to choose your tuning parameters.

```{r}
library(e1071)
#interaction x1 * x2
testing <- data.frame(X1, X2, X1*X2, Y=as.factor(Y))
training <- data.frame(X1, X2, X1*X2, Y=as.factor(Y))
#view(testing)

fit.svm <- svm(Y~.,data=training,kernel="linear",cost=1)

tune.out <- tune(svm,Y~.,data=training,kernel="linear",
  ranges=list(cost=seq(0.01,5,length.out=100)))
fit.svc <- tune.out$best.model

fit.svc
dim(fit.svc$coefs)[1]/dim(training)[1] # percent of data used as support vectors
```

Describe what the SVM is doing, i.e., how it classifies points.

The SVM is creating a hyperplane using the 2 predictor x1 and x2 variables, as well as the interaction term x1*x2. It looks to perfectly separate the points such that $$\beta_0 +\beta_1x_1+\beta_2x_2+\beta_3x_1x_2=0$$

Which looks to define anything above the plane as a 1, and anything below the plane as a -1. The interaction term helps to define where to place the plane such that it gets a more accurate prediction.

Calculate the sensitivity, specificity, and overall error rate (misclassification rate) for your SVM given the test set of 100 data points.

```{r}
#view(testing)
tab <- table(true=testing[,"Y"],pred=predict(fit.svc,newdata=testing))
tab
#sensitivity
tab[2,2]/(tab[2,1]+tab[2,2])
#specificity
tab[1,1]/(tab[1,1]+tab[1,2])
#Error rate
(tab[1,2]+tab[2,1])/(tab[1,2]+tab[2,1]+tab[1,1]+tab[2,2])
```
Make a prediction for a new point (x1, x2) = (0.25, 0.25) or (0.25, 0.75) or (0.75, 0.25) or (0.75, 0.75) for each fitted model. Each individual will have a different point for their predictions. Use the same one you used for Lab1

```{r}
#Point prediction (.75, .25)
point <- data.frame(X1=c(.75), X2=c(.25), X1...X2=c(.25*.75))

pointpred<-predict(fit.svc, point)
pointpred
```
The SVM predicts the point (.75, .25, .25*.75) is a 0.

Comment on how the prediction differs from your prediction(s) in Lab1 and, given this new prediction, describe what actions (Q3) you recommend given your Q2 results.

The model predicted a 0 in both cases. Given the new interaction term and the Q2 results, I would say its more likely than not that for this specific point it actually is a 0. However the model had a fairly high error rate at ~41.3%. So ethically there may be complications as the model under performs what I would be comfortable with implementing in circumstances where error leads to largely damaging outcomes. I would recommend gathering more data and predictors to lower this error rate in those circumstances. 


#### What teams need to do

1.  Create a new team SVM that predicts the test set better than any individual SVM.

```{r}
#create poly svm and do a lot of tuning

#create radial svm with interaction
set.seed(123)
trainingteam <- data.frame(
  X1 = training$X1,
  X2 = training$X2,
  X1X2 = training$X1 * training$X2,
  X1_sq = training$X1^2,
  X2_sq = training$X2^2,
  Y = as.factor(training$Y)
)

testingteam <- data.frame(
  X1 = testing$X1,
  X2 = testing$X2,
  X1X2 = testing$X1 * testing$X2,
  X1_sq = testing$X1^2,
  X2_sq = testing$X2^2,
  Y = as.factor(testing$Y)
)
set.seed(123)
teamtune <- tune(svm, Y ~ .,
                 data = trainingteam,
                 kernel = "radial",
                 scale=TRUE,
                 ranges = list(
                   #degree=c(2,3),
                   cost = c(0.01, 0.1, 1, 10, 100),
                   gamma = c(0.001, 0.1, 0.25, 0.5, 1),
                 #  coef0 = c(0, 1)),
                 tunecontrol = tune.control(cross = 5)))
summary(teamtune)


bestteam <- teamtune$best.model
#summary(bestteam)

teampred <- predict(bestteam, testingteam)
teammatrix <- confusionMatrix(data=teampred,
                                reference=testingteam$Y)
teammatrix

#does ever so slightly better with an accuracy of 0.593

#### Team Plot

```{r}
tb <- data.frame(Model = c("Linear","Polynomial:2", "Interaction","Team SVM"),
                 Overall_Error = c(0.49,0.5,.41,0.407),
                 Sensitivity = c(1,0.74,.83,.821),
                 Specificity = c(0,0.26,.31,.331))

library(tidyr)
library(dplyr)
library(ggplot2)

tb_long <- tb %>%
  pivot_longer(
    cols = -Model,
    names_to = "Metric",
    values_to = "Value"
  )

ggplot(tb_long, aes(x = Value, y = Model, color=Metric)) +
  geom_point(size = 3) +
  facet_wrap(~Metric, scales = "free_x") +
  scale_color_manual(values = c(
    "Error Rate" = "red",
    "Sensitivity" = "blue",
    "Specificity" = "darkgreen"
  ))+
  theme_minimal()

```

### Team Function
```{r}

#Create Function to 
crows_func <- function(seed,testing){
  set.seed(seed)
  
  training <-as.data.frame(training)
  testing <- as.data.frame(testing)
  
  trainingteam <- data.frame(
  X1 = training$X1,
  X2 = training$X2,
  X1X2 = training$X1 * training$X2,
  X1_sq = training$X1^2,
  X2_sq = training$X2^2,
  Y = as.factor(training$Y)
)
  

testingteam <- data.frame(
  X1 = testing$X1,
  X2 = testing$X2,
  X1X2 = testing$X1 * testing$X2,
  X1_sq = testing$X1^2,
  X2_sq = testing$X2^2,
  Y = as.factor(testing$Y)
)

set.seed(seed)
teamtune <- tune(svm, Y ~ .,
                 data = trainingteam,
                 kernel = "radial",
                 scale=TRUE,
                 ranges = list(
                   #degree=c(2,3),
                   cost = c(0.01, 0.1, 1, 10, 100),
                   gamma = c(0.001, 0.1, 0.25, 0.5, 1),
                 #  coef0 = c(0, 1)),
                 tunecontrol = tune.control(cross = 5)))
best_model <- teamtune$best.model

preds <- predict(best_model, testingteam)
  actual <- testingteam$Y
  
  TP <- sum(preds == 1 & actual == 1)
  TN <- sum(preds == 0 & actual == 0)
  FP <- sum(preds == 1 & actual == 0)
  FN <- sum(preds == 0 & actual == 1)
  
  misclass_rate <- mean(preds != actual)
  sensitivity   <- ifelse((TP + FN) == 0, NA, TP / (TP + FN))
  specificity   <- ifelse((TN + FP) == 0, NA, TN / (TN + FP))
  
  results <- data.frame(
    Misclassification_Rate = misclass_rate,
    Sensitivity = sensitivity,
    Specificity = specificity
  )
  
  return(results)
}
```

2.  Describe what the SVM is doing, i.e., how it classifies points.
The team SVM uses a radial kernel, and tunes the parameters using 5 fold cross validation to choose different cost values and gamma values. The SVM using a radial kernel works by creating a hyperplane through the points in the data such that $f(x)=\beta_0+\sum^\infty_i\alpha_ik(x,x_i)$ . Where it predicts 1 if $f(x)>0$, or -1 otherwise. $k(x,x_i)=\exp(-\gamma(||x-x_i||^2))$ and this is where our cross validation can fine tune all of these metrics to find the best model. The equation basically checks how close different points are to the support vector. Instead of having to add interaction terms manually, the kernel takes into account many with the radial kernel. The gamma parameter tells the model how much they should weight or pay attention to the support vectors. The cost parameter tells the model how much it should sacrifice when it comes to placing a boundary around points.

4.  Provide R code that can easily be run in class to test the performance of your team's SVM (on a new test set). The code must accept a test dataset of 100 X and y values and output the misclassification rate, sensitivity, and specificity of the SVM. More specifically, write the code so that the only thing that needs to change is the random seed. Add some aspect of your team name to your code so we can distinguish between different teams' functions. Email your .qmd file to the professor. (Upload a pdf to Canvas.) The team with the best classifier on the test set the professor will generate in class (using a different seed) will get 10 bonus points. Second place will get 5 bonus points. Third place will get 4 bonus points. Fourth place will get 3 bonus points. Fifth place will get 2 bonus points. Sixth place will get 1 bonus point. Only teams whose code runs (in class) on the first try will get any bonus points.

5.  In the team section of this lab, write a paragraph about individual contributions. Each individual must comment on their contributions to the lab. Only individuals who contribute to the team section will get points for the team section.

Corva created the team testing set and created the team svm
Averey added the explanation of what the SVM is doing, and how it classifies points.
Parker created the visualization for the svms
#### **Some intended outcomes from this assignment:**

-   You will individually get practice fitting a SVM
-   You will compare the predictive performance of multiple SVMs and previously used classification methods
-   You will make predictions based on your SVM
-   You will compare models visually
-   You will gain experience collaborating with your teammates on an applied problem
