---
title: "Lab4: Clustering (and PCA)"
author: "STAT 4610/5610"
format: pdf
editor: visual
date: 2025-04-25
---

## **Due: 11:59PM Monday, May 5 on Canvas as a knitted pdf file of your team's Quarto document**

1.  You will individually create clusters from the `Attitude` data set we've worked with in class. You will create clusters of the data points (the respondents) AND also clusters of the attitudes.
2.  You will individually interpret the clusters of the data points and compare them to the established `Group` of respondents.
3.  As a team, you will decide the best way to cluster this data set. You will interpret these clusters.

### The Context

Four groups of individuals responded to a survey about Attitudes in Collaboration conducted by Prof. Vance and Laily Mualifah, Dr. Vance's colleague in Indonesia (where he spent the 2023-24 academic year as a Fulbright Scholar experimenting how to teach interdisciplinary collaboration to Indonesian statistics and data science students). The goal of the research project is to see what differences in attitudes are held by Indonesian students compared to Indonesian professionals compared to students in the U.S.

If you haven't taken the survey, you still can (for 3 bonus points) at https://forms.gle/Q7d5XVMeub5Y2DLN6. (Note: if you already took the survey you've already been given 5 bonus points. Don't take the survey a second time. If you do, your bonus score will change from 5 to 3.)

#### The Survey

The survey asks respondents to rate how often they have an attitude and then to what degree they think that attitude is detrimental to or beneficial for interdisciplinary collaboration. Then some demographic and experience questions are asked at the end of the survey.

The four groups who took the survey were:

-   Indonesian professionals (not undergraduate students) in 2024, labeled `IndoProf24`

-   Indonesian undergraduates in 2024 after taking a course on Statistical Collaboration, labeled `IndoPost24`

-   Students fromStatistical Collaboration at CU Boulder. Both ugrads and grad students, labeled `StatCollab`

-   Students from Statistical Learning at CU Boulder in Spring 2025, labeled `SLSp25`

#### The Data

The data can be downloaded from: <https://drive.google.com/file/d/179nQH4lO2oUR_9l8sySFKZI7RDPU5rMG/view?usp=sharing>

The questions were asked in groups of five. The first five questions were about how often the respondent had Attitudes 1-5, then how beneficial/detrimental Attitudes 1-5 would be for collaboration, then how often for Attitudes 6-10 and how beneficial/detrimental for Attitudes 6-10, and so on until Attitudes 21-25.

The column name "A1.freq" indicates Frequency of Attitude 1 (How often do you have this attitude?) on a scale from 1 (Never) to 5 (Always). The ratings numbers 1-5 correspond to the ratings "Never" "Rarely" "Sometimes" "Usually" and "Always" (and their Indonesian translations) for the Frequency questions.

The column name "A1.fav" indicates how detrimental/beneficial the respondent thinks that attitude is on a six-point scale: "Extremely Detrimental (1)" "Detrimental (2)" "Somewhat Detrimental (3)", "Somewhat Beneficial (4)", "Beneficial (5)", and "Extremely Beneficial (6)" (and their Indonesian translations).

The 25 attitudes are downloadable here: <https://docs.google.com/document/d/18Hjt4LPC2vkko9Bxm_7NqVgKYLEto8Mz/edit?usp=sharing&ouid=108624562691464780001&rtpof=true&sd=true>

The `Group` variable in the data set has values "StatCollab", "SLSp25", "IndoProf24", and "IndoPost24" corresponding to American students in Statistical Collaboration F24/Sp25, students in Statistical Learning Sp25, Indonesian professionals in 2024, and Indonesian undergraduates in 2024 after taking a course on Statistical Collaboration.

### Instructions for Lab4

#### What individuals need to do (20 pts)

1.  Implement K-means clustering on the data points. Use a sufficient number of starting values (20 is probably fine). Choose $K\in \{2,3,4,5,6\}$. Each individual should choose a different $K$.
2.  Interpret your results. To what degree do your $K$ clusters correspond to the four groups?
3.  For the same $K$, cluster the attitudes (cluster the transpose of the $X$ matrix). Interpret your results. Comment on the patterns you see within the clusters of attitudes.
4.  Use hierarchical clustering to cluster the data points. Each individual should use a different linkage (complete, single, average, or centroid). You must try at least two different distance measures (e.g., Euclidean, Manhattan, correlation). Comment on how many clusters you chose and why? Comment on which distance measure you used for your final clusters and why.
5.  Plot the "best" dendrogram. Interpret your results. To what degree do your $K$ clusters correspond to the four groups?
6.  Cluster the attitudes using hierarchical clustering. Use a different linkage (or use more than one and pick the best one). Justify your choice of distance measure.
7.  Plot the "best" dendrogram. Interpret your results. Compare the patterns you see within the hierarchical clusters of attitudes with the patterns you saw using K-means.

### Parker Individual Section

```{r,message=FALSE,warning=FALSE}
data <- read_csv("Attitude.clean.csv")

```
### Parker Individual Section
```{r}
#PCA to Visualize Groups

pca <- prcomp(data[,-1],scale=TRUE)

data$Group <- as.factor(data$Group)
plot(pca$x[,1:2],pch=19,col=as.factor(data$Group), main="PC1 and PC2")
legend("topright",legend=levels(data$Group),pch=19,col=1:6)
```
### K-Means Clustering
I scaled the data to have mean zero and standard deviation one because some questions range from 1-5, while other range from 1-6. 
#### (1)
```{r}
set.seed(123)

mat_dat <- as.matrix(data[,-1])
mat_dat <- scale(mat_dat)

cluster_4 <- kmeans(mat_dat, 4, nstart=20)
cluster_4$tot.withinss
```
```{r}
#Visualize 4 Clusters Via PCA

plot(pca$x[,1:2],pch=19,col=as.factor(cluster_4$cluster), 
     main="PC1 and PC2 With 4-Means Cluster Labels")
```
#### (2)
If we compare the plot of the first two principle components with the actual group labels to the plot with the labels obtained from the 4-means clustering, then it is clear that 4 groups do not represent the data that well. The original data seems to be much more mixed compared to the 4-means clustering.

#### (3)
```{r}
set.seed(123)
mat_dat_t <- t(mat_dat)
mat_dat_t <- scale(mat_dat_t)

trans_cluster <- kmeans(mat_dat_t,4,nstart=20)
trans_cluster$tot.withinss
```
```{r}
#Plot PC1 and PC2 with labels obtained from 4-means clustering of 
#transposed x matrix
pca_t <- prcomp(t(data[,-1]),scale=TRUE)

col_groups <- as.factor(colnames(data[,-1]))

plot(pca_t$x[,1:2],pch=19,col=col_groups,main="PC1 and PC2 of Transpose")
legend("topright",legend=levels(col_groups),pch=19,col=1:6)
```
```{r}
#PC1 and PC2 with 4-means cluster labels 
plot(pca_t$x[,1:2],pch=19,col=trans_cluster$cluster,main="PC1 and PC2 of Transpose: 4-Means Labels")
```
Again, the 4-means cluster does not match the data super well. There aren't really any distinct groups if we look at the actual labels. But we are trying to group all of the different attitudes into four groups. Compared to the non-transposed data, these four groups from the k-means algorithm are more distinct. 

### Hierarchical Clustering 
####(4)

```{r}
#Hierarchical Clustering with Euclidean Measure

hc_eucl <- hclust(dist(mat_dat), method="average")

plot(hc_eucl, main="Euclidean Measure and Average Linkange",ylab="",sub="",,xlab="",
     cex=0.9)
```
#### (5)
```{r}
dd <- as.dist(1-cor(t(mat_dat)))

x <- 1:140
y<- rep(0.98,140)
hc_corr <- hclust(dd,method="average")
plot(hclust(dd,method="average"),main="Correlation Measure and Average Linkage",
     xlab="",ylab="",sub="")
lines(x,y,lty="dashed",col="red",lwd=2)
```
I think it makes the most sense to use the correlation distance measure because it can capture the similarity of attitude patterns versus just absolute differences. The correlation measure can capture patterns of individuals answering in similar ways, but not exactly the same: people who tend to give high answers and people who tend to give small answers. After observing the tree, I want to cut the tree at 0.98 to obtain 4 groups. I chose 0.98 because the vertical distance of the connecting nodes seem to yield farily distinct groups.
```{r}
#Cut Tree
corr_cut <- cutree(hc_corr,4)

#Plot PC1 and PC2 with Corr_cut labels
plot(pca$x[,1:2],pch=19,col=as.factor(corr_cut), 
     main="PC1 and PC2: Hierarchical,Correlation Measure Labels")
```
This seems to have performed slightly better than the 4-means cluster because it captures more that the points are more mixed. The two smallest groups are the least distinct. I think the correlation distance measure with the hierarchichal clustering captured a relationship in the data that the 4-means clustering didn't.

##### (6)
```{r}
#Hierarchical Clustering on transpose with Complete linkage, Euclidean Distance
ddt <- as.dist(1-cor(t(data[,-1])))

x <- 1:180
y <- rep(.85,180)
hc_trans_corr <- hclust(ddt,method="complete")
plot(hc_trans_corr,xlab="",ylab="",main="Attitudes and Complete Linkage",sub="")
lines(x,y,col="red",lty="dashed",lwd=2)
```
When we transpose the data we are trying to group together the attitudes. So I also think that correlation distance will be an appropriate measure to capture similarity in attitudes among all respondents. I think that cutting at 0.85 makes sense since the vertical distances seem to be large enough to show distinct groups. This yields 6 groups.

#### (7)

```{r}
#Cut tree and plot groups on first two principle components
trans_cut <- cutree(hc_trans_corr,6)

plot(pca_t$x[,1:2],pch=19,col=as.factor(trans_cut),main="PC1 and PC2 of Transpose: HC Groups with Correlation Measure")
```
If I compare this to the original labels, then they are clearly not mixed around enough. And If I compare this to the plot with the 4-means cluster labels, then they look relatively similar, except these groups are slightly more mixed about, which is better. 


```{r}
hc_corr2 <- hclust(dd,method="complete")
corr_cut2 <- cutree(hc_corr2,6)

plot(pca$x[,1:2],pch=19,col=as.factor(corr_cut2), 
     main="PC1 and PC2: Hierarchical,Correlation Measure Labels and Complete Link")
``` 

### Averey Copeland Individual Section
1)
```{r}
library(ISLR2)
library(tidyverse)
set.seed(17)
data <- read.csv("Attitude.clean.csv")
#View(data)

#k2 <- km.out(data, 2, nstart = 20)
X <- data[,-1] #Excluding the groups
dim(X)
```
2) Interpret your results. To what degree do your $K$ clusters correspond to the four groups?

```{r}

table(km.out$cluster,data$Group)
```
This table shows that it clustered group 1 with primarily IndoPost24, IndoPre25, and IndoProf24. Group 2 had a mix of those groups plus the majority of the classes in the United States. There must be an underlying pattern in which people responded in group 1 that was different from the United States answers.
3)
```{r}
transpose <- t(X)
km.transpose <- kmeans(transpose, 2, nstart = 20)

#table(km.transpose$cluster,rownames(transpose))
pca<- prcomp(transpose, scale = TRUE)
plot(pca$x[,1], pca$x[,2],pch=19,col=km.transpose$cluster,main="PC of Transpose")
```
4)
```{r}
# Creating distance measures
cor_spearman <- cor(t(X), method = "spearman") # Transpose so the rows are the attitude variables
dist_spearman <- as.dist(1 - cor_spearman)


# Using hclust
hc.complete <- hclust(dist_spearman, method = "single")
plot(hc.complete)
cut1<-cutree(hc.complete, k=2)
#plot(cut1)

dist_euclidean <- dist(X, method = "euclidean")
hc.complete2 <- hclust(dist_euclidean, method = "single")
#plot(hc.complete2)
#cutree(hc.complete, k=2)
cut2<-cutree(hc.complete, k=5)
#plot(cut2)
```
5)
```{r}
plot(hc.complete)
```
Again, the clusters using the single method are not good. I can't find a place to cut since the clusters all merge gradually, and there is no obvious point of change. I would say this corresponds very little.
6)
```{r}
cor_pearson <- cor(t(X), method = "pearson") 
dist_pearson <- as.dist(1 - cor_pearson)
hc.complete2 <- hclust(dist_pearson, method = "average")
plot(hc.complete2)
cut <- cutree(hc.complete2, k=3)
plot(hc.complete2, cut)
```
I chose euclidean distance since it just measures how close/far peoples answers are from one another. So that people that choose relatively the same answers overall then they will have a smaller distance from one another. This highlights people that choose 1 then 6 vs someone who chooses 1 then 2 for example. Since they are closer in how they would answer their questions euclidean distance would capture that.
7)
```{r}
plot(hc.complete2)
```
The dendrogram shows that most peoples answers were similar, since overall, the majority of the merges happen at around .35. When doing cuts at k=2 and 3 it showed that there is 1 person that answered differently from others in the data set as it consistently grouped a single person in its own cluster with both of these cuts. This is similar to the k means pattern since it also had a majority of observations grouped together with a minority of observations placed in group 1. 


## Corva Individual Part

```{r}
install.packages("tidyverse")
library(tidyverse)

att.data <- read.csv("Attitude.clean.csv")

att.data.numeric <- att.data %>%
  select(-Group)

attitudes.labels <- colnames(att.data.numeric)
groups.labels <- att.data$Group
```


K-Means for the data points:

```{r}
set.seed(123)
km.six <- kmeans(att.data.numeric, 6, nstart=20)
table(km.six$cluster, groups.labels)
```

My six clusters don't do the best job of corresponding to the *six* groups. It appears to pick up on bigger trends, like cluster 3 being mostly SLS/ Stat Collab or Cluster 5 being some sort of Indonesian group. 


K-Means for Attitudes: 

```{r}
set.seed(123)
km.six.att <- kmeans(t(att.data.numeric), 6, nstart=20)
table(km.six.att$cluster, attitudes.labels)

km.six.att$cluster
attitudes.labels
# table(km.six.att$cluster, attitudes.labels)
```

Similar attitudes seem to be in the same cluster, for example, attitudes 7 and 12: "Usually, domain experts do want to learn new statistics; they just want to take the easiest path/ Domain experts only need to know the statistical analysis results; they don't need to understand the statistical methods used". Based on the responses about attitude frequency/ benefits, clustering grouped similar attitudes without even knowing what the specific attitudes were. In addition, it appears that frequency and how beneficial people find their attitudes seem to be in the same cluster for most attitudes. 


Hierarchical clustering for the data points:

```{r}
#scale the variables
att.data.numeric.2 <- scale(att.data.numeric)
km.six.att.comp <- hclust(dist(att.data.numeric.2), method="complete")
km.six.att.man <- hclust(dist(att.data.numeric.2, method="manhattan"), method="complete")

#plotting it
pcac <- prcomp(att.data.numeric.2, scale. = TRUE)
plot(pcac$x[,1:2],
     pch = 19,
     col = as.factor(c.best.hier),
     main = "PC1 and PC2: Hierarchical (Complete Linkage, Euclidean Distance) Clusters")

#plot(km.six.att.comp, main="complete linkage, euclidean distance")
#plot(km.six.att.man, main="complete linkage, manhattan distance")
```

For my final clusters, I used the Euclidean distance because separation between the groups in the plotted dendrogram appeared to be more distinct. Manhattan distance didn't seem appropriate because the data points weren't laid out in a grid-like fashion. If I didn't know anything about the data, I would choose four clusters because cutting the tree for 6 or 5 clusters results in only one observation being in group 6/5. Cutting the tree at around height 15 cuts at a place with long vertical lines, and results in four distinct groups. However, I'll chose 6 clusters because I know there are six groups in the data.

```{r}
c.best.hier <- cutree(km.six.att.comp, 6)
plot(km.six.att.comp, main="complete linkage, euclidean distance")
c.best.hier

table(c.best.hier, groups.labels)
```

The 6 clusters don't very well correspond to the six groups. Just like the K-means clustering, the clustering seemed to pick up on larger groups, for example, putting almost all the observations for SLSp25 and SLSp26 in the same group. However, the clustering didn't pick up on smaller distinctions between the data, for example, between the Stat Learning Spring 2025 vs 2026. It's possible that the data might just not be different enough between these smaller groups.

Clustering attitudes with hierarchical clustering: 

```{r}
att.data.numeric.t <- scale(t(att.data.numeric))
km.six.att.compt <- hclust(dist(t(att.data.numeric.2)), method="average")
km.six.att.mant <- hclust(dist(t(att.data.numeric.2), method="manhattan"), method="complete")
```

For my final clusters, I used the Euclidean distance because separation between the groups in the plotted dendrogram appeared to be more distinct. Manhattan distance didn't seem appropriate because the data points weren't laid out in a grid-like fashion. I'll chose 25 clusters because I know there are 25 attitudes in the data.

It appears that the hierarchical clusters of attitudes do a good job of separating the attitudes within clusters: it appears that frequency and how beneficial people find their attitudes seem to be in the same cluster for most attitudes. It appears that the hierarchical clustering placed attitudes 6, 25, and 23 in the "wrong" clusters. The hierarchical clustering performed much better than K-means clustering, being able to pick out individual attitudes instead of finding clusters of attitudes that were somewhat similar.

```{r}
c.best.hiert <- cutree(km.six.att.compt, 25)
#plot(km.six.att.compt, main="average linkage, euclidean distance")
#c.best.hiert

#table(c.best.hiert, attitudes.labels)

#plotting it
pcat <- prcomp(att.data.numeric.t, scale. = FALSE)
plot(pcat$x[,1:2],
     pch = 19,
     col = as.factor(c.best.hier),
     main = "Hierarchical Complete Linkage, Euclidean Distance for Transpose")
```

It appears that the hierarchical clusters of attitudes do a good job of separating the attitudes within clusters: it appears that frequency and how beneficial people find their attitudes seem to be in the same cluster for most attitudes. It appears that the hierarchical clustering placed attitudes 6, 25, and 23 in the "wrong" clusters. The hierarchical clustering performed much better than K-means clustering, being able to pick out individual attitudes instead of finding clusters of attitudes that were somewhat similar.


#### What teams need to do

1.  Review and combine individual results. Which methods for clustering the data points seemed to correspond best with the pre-established groups? What $K$ gave you the best results? What linkage? What distance metric?
Hierarchical clustering yielded the best results over K-Means Clustering. The best number of clusters was six, which makes sense since there are six groups present in the data. The best linkage was the average linkage with the correlation measure. The correlation measure made sense in this context because it captures patterns of individuals who tend to give high responses together and how other individuals tend to give low responses together. 
2.  Compare the results from Principal Component Analysis (PCA) with the results of clustering the attitudes. Which is more interpretable? What is your interpretation?

The clustering of the attitudes seems more interpratable since it directly shows the clusters of how close people's attitudes were to each other. This shows patterns within the data that can be more easily interpreted. My interpretation of the clustering is that most people have similar opinions in their answers, and those that have different opinions than the majority also answered similarly within their own clusters for the most part.

3.  Create your team's final, best clustering of the data points. How well do the clusters correspond to the four groups?

```{r}
#using average linkage
#using correlation
dist_corr <- as.dist(1 - cor(t(att.data.numeric.2)))
team_hier <- hclust(dist_corr, method="average")
team_cut <- cutree(team_hier, k = 6)
#plotting it
pcat <- prcomp(att.data.numeric.2, scale. = TRUE)
plot(pcat$x[,1:2],
     pch = 19,
     col = as.factor(team_cut),
     main = "average linkage, correlation")
```
It appears that the best clustering of the data points was with average linkage and correlation distance meausre. The clusters correspond mostly well to the six groups, with most of the groups in the PCA plot being distinct from each other.
4.  Create your team's final, best clustering of the attitudes. What patterns do you see within the clusters of attitudes?
5.  Write a paragraph about individual contributions. Each individual must comment on their contributions to the lab. Only individuals who contribute to the team section will get points for the team section. An example contribution could be: "I synthesized all of the individual results to find that such and such gave the best and most interpretable results. Then I recommended teammate X use this linkage and that distance measure for our team's cluster." Or, "I did the comparison of PCA with our clustering of the attitudes." Or, "I created our team's best clustering of the data points based on teammate Y's synthesis of our individual results." Or, "I took everyone's code and put it together into the qmd file and knitted it." Or, "I made the plot of our team's best clusters."

#### Some intended outcomes from this assignment:

1.  You will individually get practice using K-means and hierarchical clustering
2.  You will practice using different distance measures and linkages for hierarchical clustering
3.  You will gain experience interpreting the outputs of clustering algorithms
4.  You will gain experience collaborating with your teammates on an applied problem

#### Potentially useful code below

```{r include=F}
suppressPackageStartupMessages(library(tidyverse))
dat <- read_csv("Attitude.clean.csv")
dim(dat)

X <- dat[,-1] #Excluding the groups
dim(X)

# Creating distance measures
cor_spearman <- cor(t(X), method = "spearman") # Transpose so the rows are the attitude variables
dist_spearman <- as.dist(1 - cor_spearman)

cor_pearson <- cor(t(X), method = "pearson") 
dist_pearson <- as.dist(1 - cor_pearson)

# Using Manhattan or Euclidean distance
dist_manhattan <- dist(X, method = "manhattan") #Manhattan distance counts 
# the "blocks" between data points as if you were walking along the streets of 
# Manhattan (i.e., you can't cut diagonally through a city block)
dist_euclidean <- dist(X, method = "euclidean") #Euclidean distance is the 
# diagonal "shortest path" between points A and B.

# Using hclust
hc.complete <- hclust(dist_spearman, method = "complete")
cutree(hc.complete, k=4)

# Kmeans
km.out <- kmeans(X, k=4, nstart = 20)

# Doing PCA
pr.out <- prcomp(X,scale=F) #The attitudes are basically already on the same scale, so scaling the attitudes would confuse things, though A1.freq is on a 5-point scale and A1.fav is on a 6-point scale

# Find means of the four groups
SL.means <- colMeans(dat[dat[,1]=="SLSp25",2:51])
StatCollab.means <- colMeans(dat[dat[,1]=="StatCollab",2:51])
IndoProf.means <- colMeans(dat[dat[,1]=="IndoProf24",2:51])
IndoPost.means <- colMeans(dat[dat[,1]=="IndoPost24",2:51])

# Then transform the group centroids into the Principal Component space
C.matrix <- rbind(StatCollab.means,SL.means,IndoProf.means,IndoPost.means)

#Transform C.matrix into scores using the rotation matrix of the first two principal components
W2 <- pr.out$rotation[,1:2]
Group.scores <- C.matrix %*% W2
Group.scores <- mutate(as.data.frame(Group.scores),groups=rownames(Group.scores))
ggplot(data=Group.scores, aes(x=PC1,y=PC2,col=groups)) +
  geom_point()

# Comparing found clusters with groups
# What percentage of each cluster is each group?
table(km.out$cluster,dat$Group) #can transform this into percentages if necessary

# We can do the same with clusters found from hiearchical clustering
table(cutree(hc.complete,k=4),dat$Group) #These clusters aren't very good.

```
